Think deep for reef resilience

### 2025-07-02 paper/paper_draft.qmd
Remove management area information and add carbonate budget text

### Reef health metrics

ADRIA-CoralBlox was run for 4096 *sampled* scenarios per GCM. A median absolute coral cover timeseries for each reef was calculated for each GCM. For use in timeseries clustering reef cover timeseries were scaled by the total area of the reef to represent timeseries as a percentage of total reef area (range 0 - 100%). The proportion of each functional group occupying the reef (taxa evenness) was also investigated as a metric for timeseries analysis, but was found to not sufficiently capture the declining condition of reefs suffering from very low absolute coral cover.

@@

Timeseries analysis is performed by first grouping reefs into three different scales, GBRMPA reefal bioregion, GBRMPA management area and GBR-wide, to investigate the consistency of cluster identification. To identify groups of reefs at each analysis scale that are resilient, intermediate and non-resilient we clustered reef coral cover timeseries into three clusters at each scale. To quantify the behaviour of timeseries and identify timeseries that behave similarly we calculate the complexity-invariant distance (CID) between each timeseries. The complexity-invariant distance refers to the Euclidean distance between two timeseries multiplied (corrected) by the ratio of the complexities of the timeseries [@batista2014]. These values create a distance matrix of n * n dimensions (where n is the number of timeseries in the group) that is clustered using a k-medoids clustering algorithm [@steinmann2020].

@@

To maintain coral cover over time coral reefs require a threshold of coral cover known as a carbonate budget. Estimates of the carbonate budget required to maintain a healthy coral reef range from 10 to 20% of a reef's total area. Using median absolute cover timeseries for each reef we calculated the number of years each reef maintains coral cover above a carbonate budget threshold level. Threshold values from 10 to 20% (inclusive) were used in analyses.

### Further analyses

@@

@@

## Reef carbonate budget
The number of years reefs maintain coral cover above a required carbonate budget level substantially decreases as the required carbonate budget threshold (as a percentage of reef area) increases. Reefs with greater median depth levels maintain coral cover above required carbonate budget levels for longer than reefs at shallow depths [@fig-access-esm1-5-depth-carbonate]. There are no clear relationships between reef total connectivity levels and the number of years reefs maintain coral cover above carbonate budget threshold levels [@sfig-access-esm1-5-conn-carbonate].

![Number of years reef coral cover exceeds the carbonate budget threshold level from 2025 to 2099. The carbonate budget thresholds assessed range from 10 to 20% of total reef area. Point colour represents the median reef depth level in metres. Displays results from median timeseries using ACCESS-ESM1-5.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/depth_carbonate_scatter.png){#fig-access-esm1-5-depth-carbonate width=`{julia} carb_width`}

@@

Across most bioregions there are not consistent differences between clusters in connectivity metrics such as source-to-sink connectivity ratio (@sfig-access-esm1-5-bioregion-logsosi; @sfig-access-cm2-bioregion-logsosi; @sfig-ec-earth3-veg-bioregion-logsosi; @sfig-gfdl-cm4-bioregion-logsosi; @sfig-noresm2-mm-bioregion-logsosi) and total connectivity strength (@sfig-access-esm1-5-bioregion-totalconnectivity; @sfig-access-cm2-bioregion-totalconnectivity; @sfig-ec-earth3-veg-bioregion-totalconnectivity; @sfig-gfdl-cm4-bioregion-totalconnectivity; @sfig-noresm2-mm-bioregion-totalconnectivity).
<!-- 
## Management area scale clustering

Within the four management areas, timeseries clustering analysis identified three clusters of reefs using relative coral cover timeseries. These clusters form groups of reefs with ‘low, medium and high’ relative coral cover, categorised by their median coral cover between 2030 and 2040. The ‘high’ cover cluster of reefs contains reefs that have continuing recovery of corals throughout the future climate scenario. The medium and low clusters contain many reefs that decrease to half their initial coral cover levels and largely do not display the levels of recurring recovery that reefs in high clusters experience (@sfig-access-esm1-5-manarea-timeseries; @sfig-access-cm2-manarea-timeseries; @sfig-ec-earth3-veg-manarea-timeseries; @sfig-gfdl-cm4-manarea-timeseries; @sfig-noresm2-mm-manarea-timeseries).

Within each bioregion reef median depth level increases with increasing reef relative coral cover. Reefs in the low cluster have the lowest median depth levels, while reefs in the high cluster have the highest median depth levels within each management area (@sfig-access-esm1-5-manarea-depth; @sfig-access-cm2-manarea-depth; @sfig-ec-earth3-veg-manarea-depth; @sfig-gfdl-cm4-manarea-depth; @sfig-noresm2-mm-manarea-depth).

Correlation between reef total coral cover and the DHW levels they experience throughout the time period increases as coral relative cover increases, with reefs in the high cluster having the highest median total cover – DHW correlation (@sfig-access-esm1-5-manarea-dhwcor; @sfig-access-cm2-manarea-dhwcor; @sfig-ec-earth3-veg-manarea-dhwcor; @sfig-gfdl-cm4-manarea-dhwcor; @sfig-noresm2-mm-manarea-dhwcor).

However, reefs with higher relative coral cover levels do not experience lower mean DHW levels throughout their timeseries. There are no consistent substantial differences in mean DHW levels across the identified timeseries clusters (@sfig-access-esm1-5-manarea-dhwmean; @sfig-access-cm2-manarea-dhwmean; @sfig-ec-earth3-veg-manarea-dhwmean; @sfig-gfdl-cm4-manarea-dhwmean; @sfig-noresm2-mm-manarea-dhwmean).

Within the Far Northern management area ‘low’ cluster reefs have substantially lower median DHW – tolerance levels than ‘medium’ and ‘high’ clusters, while there are no substantial differences in DHW – tolerance between clusters in other management areas (@sfig-access-esm1-5-manarea-dhwtol; @sfig-access-cm2-manarea-dhwtol; @sfig-ec-earth3-veg-manarea-dhwtol; @sfig-gfdl-cm4-manarea-dhwtol; @sfig-noresm2-mm-manarea-dhwtol).

Like clusters identified within bioregions, within the Mackay-Capricorn, Far Northern and Cairns-Cooktown management areas the initial proportion of coral-habitable area initially occupied by corals decreases in clusters with higher relative coral cover (@sfig-access-esm1-5-manarea-initialproportion; @sfig-access-cm2-manarea-initialproportion; @sfig-ec-earth3-veg-manarea-initialproportion; @sfig-gfdl-cm4-manarea-initialproportion; @sfig-noresm2-mm-manarea-initialproportion).

At management area analysis scales, there are not consistent substantial differences in source – to – sink ratios (@sfig-access-esm1-5-manarea-logsosi; @sfig-access-cm2-manarea-logsosi; @sfig-ec-earth3-veg-manarea-logsosi; @sfig-gfdl-cm4-manarea-logsosi; @sfig-noresm2-mm-manarea-logsosi) and total connectivity strength (@sfig-access-esm1-5-manarea-totalconnectivity; @sfig-access-cm2-manarea-totalconnectivity; @sfig-ec-earth3-veg-manarea-totalconnectivity; @sfig-gfdl-cm4-manarea-totalconnectivity; @sfig-noresm2-mm-manarea-totalconnectivity).

## Great Barrier Reef scale clustering

Timeseries clustering identified three clusters in reef relative cover timeseries for the whole GBR scale. A ‘low’ cluster of reefs with little recovery following initial declines, a medium cluster of reefs with median relative cover below 0.5 post 2028, and a high cluster of reefs with median relative cover above 0.5 until 2050 (@sfig-access-esm1-5-gbr-timeseries; @sfig-access-cm2-gbr-timeseries; @sfig-ec-earth3-veg-gbr-timeseries; @sfig-gfdl-cm4-gbr-timeseries; @sfig-noresm2-mm-gbr-timeseries).

The clusters identified with timeseries clustering at the GBR scale show similar patterns to clusters identified at bioregion and management area scales. Reef depth (@sfig-access-esm1-5-gbr-depth; @sfig-access-cm2-gbr-depth; @sfig-ec-earth3-veg-gbr-depth; @sfig-gfdl-cm4-gbr-depth; @sfig-noresm2-mm-gbr-depth) and total cover – DHW correlation (@sfig-access-esm1-5-gbr-dhwcor; @sfig-access-cm2-gbr-dhwcor; @sfig-ec-earth3-veg-gbr-dhwcor; @sfig-gfdl-cm4-gbr-dhwcor; @sfig-noresm2-mm-gbr-dhwcor) decrease, while the proportion of coral-habitable area initially occupied by corals increases (@sfig-access-esm1-5-gbr-initialproportion; @sfig-access-cm2-gbr-initialproportion; @sfig-ec-earth3-veg-gbr-initialproportion; @sfig-gfdl-cm4-gbr-initialproportion; @sfig-noresm2-mm-gbr-initialproportion), as the clusters decrease from ‘high’ to ‘low’ cover levels. There are no substantial differences between the clusters in mean DHW (@sfig-access-esm1-5-gbr-dhwmean; @sfig-access-cm2-gbr-dhwmean; @sfig-ec-earth3-veg-gbr-dhwmean; @sfig-gfdl-cm4-gbr-dhwmean; @sfig-noresm2-mm-gbr-dhwmean) and DHW tolerance (@sfig-access-esm1-5-gbr-dhwtol; @sfig-access-cm2-gbr-dhwtol; @sfig-ec-earth3-veg-gbr-dhwtol; @sfig-gfdl-cm4-gbr-dhwtol; @sfig-noresm2-mm-gbr-dhwtol) levels of reefs. Similarly, there are no substantial differences between clusters in connectivity metrics source – to – sink ratio and total connectivity strength (@sfig-access-esm1-5-gbr-logsosi; @sfig-access-cm2-gbr-logsosi; @sfig-ec-earth3-veg-gbr-logsosi; @sfig-gfdl-cm4-gbr-logsosi; @sfig-noresm2-mm-gbr-logsosi; @sfig-access-esm1-5-gbr-totalconnectivity; @sfig-access-cm2-gbr-totalconnectivity; @sfig-ec-earth3-veg-gbr-totalconnectivity; @sfig-gfdl-cm4-gbr-totalconnectivity; @sfig-noresm2-mm-gbr-totalconnectivity). -->

@@

Timeseries analysis has been used to identify three groups of reefs with similar declining (low), recovering (high), and intermediate (medium) coral cover trajectories throughout future scenarios. Under all GCMs most reefs suffer substantial declines in coral cover from 2025 levels. NorESM2-MM and ACCESS-ESM1-5 GCMs convey the most coral cover maintanance into the future. The differences between low, medium and high reef clusters in their timeseries are most notable under these two GCMs. At all analysis scales reef clusters show similar patterns in their reef characteristics such as depth and proportion of habitable area initially occupied.

@@

Overall, our findings do not suggest that resilient reefs in ‘high’ cover clusters have greater thermal tolerance than reefs in lower coral cover clusters. This is in contrast with literature suggesting corals have the capacity to adapt to increasing climate change pressures [@humanes2022; @lachs2024], including the investigation by Bairos-Novak et al. [-@bairos-novak2021] that revealed high heritability of bleaching and survival response traits used to inform the distributions of thermal tolerance traits used in this investigation. The lack of influence of thermal tolerance on reef outcomes is likely due to reef being impacted by overall similar levels of thermal stress, combined with the greater importance of factors such as reef depth and initial coral cover.

@@

Furthermore, our results suggest that the capacity of reefs to maintain healthy coral habitat by being above positive net carbonate budget levels is greater for reefs at deeper depths than shallow depths. While the maintanence of positive carbonate budget levels does not appear to be influenced by reef connectivity metrics. The maintanence of positive carbonate budgets is important for reef resilience under climate change.

@@

@@

![Number of years reef coral cover exceeds the carbonate budget threshold level from 2025 to 2099. The carbonate budget thresholds assessed range from 10 to 20% of total reef area. Point colour represents log scaled reef total connectivity strength. Displays results from median timeseries using ACCESS-ESM1-5.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/log_total_strength_carbonate_scatter.png){#sfig-access-esm1-5-conn-carbonate width=`{julia} carb_width`}

@@

![Log source – to – sink connectivity ratios for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines within each violin represent the median log source – to – sink ratio for each cluster and grey points represent the ratio value for each reef. Vertical grey lines represent a log source – to – sink ratio of 0 indicating equal outgoing and incoming connectivity levels. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/so_to_si_bioregion_violin.png){#sfig-access-esm1-5-bioregion-logsosi width=`{julia} violin_width`}

![Log total connectivity strength for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines within each violin represent the median log total connectivity strength for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/log_total_strength_bioregion_violin.png){#sfig-access-esm1-5-bioregion-totalconnectivity width=`{julia} violin_width`}
<!-- 
### Management area scale

![Reef relative coral cover timeseries clustered within each management area. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each management area. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each management area. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each management area. Solid lines represent the cluster median timeseries and bands represent the 2.5 to 97.5 percentile confidence intervals. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/management_area/management_area_cover_timeseries.png){#sfig-access-esm1-5-manarea-timeseries width=`{julia} timeseries_width`}

![Median depth for reefs within each management area, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each management area. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each management area. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each management area. Solid lines represent the median depth for each cluster and grey points represent the depth value for each reef. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/management_area/depth_management_area_violin.png){#sfig-access-esm1-5-manarea-depth width=`{julia} violin_width`}

![Total coral cover – DHW Pearson’s correlation for reefs within each management area, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each management area. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each management area. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each management area. Solid lines represent the median cover – DHW correlation value for each cluster and grey points represent the total cover – DHW correlation value for each reef. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/management_area/dhw_cover_cor_management_area_violin.png){#sfig-access-esm1-5-manarea-dhwcor width=`{julia} violin_width`}

![Mean timeseries DHW for reefs within each management area, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each management area. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each management area. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each management area. Solid lines represent the median mean - DHW value for each cluster and grey points represent the mean - DHW value for each reef. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/management_area/mean_dhw_management_area_violin.png){#sfig-access-esm1-5-manarea-dhwmean width=`{julia} violin_width`}

![Mean timeseries DHW – tolerance value for reefs within each management area, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each management area. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each management area. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each management area. Solid lines represent the median DHW – tolerance value for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/management_area/dhw_tolerance_management_area_violin.png){#sfig-access-esm1-5-manarea-dhwtol width=`{julia} violin_width`}

![Initial proportion of coral-habitable area occupied by coral cover for reefs within each management area, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each management area. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each management area. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each management area. Solid lines represent the median occupied proportion value for each cluster and grey points represent the value for each reef. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/management_area/initial_proportion_management_area_violin.png){#sfig-access-esm1-5-manarea-initialproportion width=`{julia} violin_width`}

![Source – to – sink connectivity ratio for reefs within each management area, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each management area. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each management area. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each management area. Solid lines represent the median source – to – sink ratio value for each cluster and grey points represent the mean source – to – sink value for each reef. Vertical grey lines represent a log source – to – sink ratio of 0 indicating equal outgoing and incoming connectivity levels. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/management_area/so_to_si_management_area_violin.png){#sfig-access-esm1-5-manarea-logsosi width=`{julia} violin_width`}

![Log total connectivity strength for reefs within each management area, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each management area. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each management area. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each management area. Solid lines within each violin represent the median log total connectivity strength for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/management_area/log_total_strength_management_area_violin.png){#sfig-access-esm1-5-manarea-totalconnectivity width=`{julia} violin_width`}

### GBR scale

![Reef relative coral cover timeseries clustered across the Great Barrier Reef. The ‘Low’ cluster (green) is the cluster with the lowest median relative cover between 2030 and 2040. The ‘Medium’ cluster (orange) is the cluster with intermediate median relative cover between 2030 and 2040. The ‘High’ cluster (blue) is the cluster with the highest median relative cover between 2030 and 2040. Solid lines represent the cluster median timeseries and bands represent the 2.5 to 97.5 percentile confidence intervals. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/gbr/gbr_cover_timeseries.png){#sfig-access-esm1-5-gbr-timeseries width=`{julia} timeseries_width`}

![Median depth for reefs, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040. Solid lines represent the median depth for each cluster and grey points represent the depth value for each reef. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/gbr/depth_gbr_violin.png){#sfig-access-esm1-5-gbr-depth width=`{julia} violin_width`}

![Total coral cover - DHW pearson's correlation for reefs, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040. Solid lines represent the median total cover – DHW correlation value for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/gbr/dhw_cover_cor_gbr_violin.png){#sfig-access-esm1-5-gbr-dhwcor width=`{julia} violin_width`}

![Initial proportion of coral-habitable area occupied by coral cover for reefs, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040. Solid lines represent the median occupied proportion value for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/gbr/initial_proportion_gbr_violin.png){#sfig-access-esm1-5-gbr-initialproportion width=`{julia} violin_width`}

![Mean DHW for reefs, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040. Solid lines represent the median DHW value for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/gbr/mean_dhw_gbr_violin.png){#sfig-access-esm1-5-gbr-dhwmean width=`{julia} violin_width`}

![Mean timeseries DHW – tolerance value for reefs, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040. Solid lines represent the median DHW - tolerance value for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/gbr/dhw_tolerance_gbr_violin.png){#sfig-access-esm1-5-gbr-dhwtol width=`{julia} violin_width`}

![Source – to – sink connectivity ratio for reefs, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040. Solid lines represent the median source – to – sink ratio value for each cluster and grey points represent the mean source – to – sink value for each reef. Vertical grey lines represent a log source – to – sink ratio of 0 indicating equal outgoing and incoming connectivity levels. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/gbr/so_to_si_gbr_violin.png){#sfig-access-esm1-5-gbr-logsosi width=`{julia} violin_width`}

![Log total connectivity strength for reefs, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040. Solid lines within each violin represent the median log total connectivity strength for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/gbr/log_total_strength_gbr_violin.png){#sfig-access-esm1-5-gbr-totalconnectivity width=`{julia} violin_width`} -->

@@

>>>>

ADRIA-CoralBlox was run for 4096 *sampled* scenarios per GCM. A median absolute coral cover timeseries for each reef was calculated for each GCM. For timeseries clustering reef cover timeseries were scaled by the total area of the reef to represent timeseries as a percentage of the total reef area. The proportion of each functional group occupying the reef (group evenness) was also investigated as a metric for timeseries analysis, but was found to not sufficiently capture the declined condition of reefs suffering from very low absolute coral cover.

@@

To identify groups of reefs at each analysis scale that are resilient, intermediate and non-resilient we clustered reef coral cover timeseries into three clusters at each scale. To quantify the behaviour of timeseries and identify timeseries that behave similarly we calculate the complexity-invariant distance (CID) between each timeseries. The complexity-invariant distance refers to the Euclidean distance between two timeseries multiplied (corrected) by the ratio of the complexities of the timeseries [@batista2014]. These values create a distance matrix of n * n dimensions (where n is the number of timeseries in the group) that is clustered using a k-medoids clustering algorithm [@steinmann2020].

Timeseries clustering analyses were performed at the reefal bioregion scale to capture groups of biogeographically similar reefs, as determined by *insert citation for bioregions*. Analyses were also performed at higher spatial scales, GBRMPA management regions and whole GBR scale. Similar results were found at large spatial scales, these can be found at *insert citation for repository*.

@@

Scleractinian corals are primary habitat structures on coral reefs. The accumulation and erosion of coral reef carbonate frameworks is controlled by a range of processes such as coral accumulation *Perry*, bioerosion *Tribollet*, physical damage generating coral rubble *Kenyon*, and calcificationa and *CaCO3* dissolution rates *Eyre 2014, 2018*. The balance of these processes and occurrance of net carbonate framework production (referred to as a positive carbonate budget) are influenced by factors such as coral cover and complexity *Perry, Desbiens*, rates of physical environmental processes and ocean acidity *Perry, Eyre*. It has been suggested that total live coral cover on a reef highly influences the maintainence of net positive carbonate budgets. Estimates of live coral cover required to maintain net positive carbonate budget levels range from 10 to 20% *Perry* of a reef's total area, with a level of 17% live coral cover for the Great Barrier Reef *Desbiens*. Using median absolute cover timeseries for each reef we calculated the number of years each reef maintains coral cover above these live coral cover threshold levels, inferring the number of years each reef maintains a net positive carbonate budget, accumulating reef framework. As in situ carbonate budgets are influenced by a range of processes, our calculations of carbonate budget maintainence are intended as estimates and do not reflect the conditions that will be experienced on a reef, but indicate relative levels of health across reefs. To capture the wide influence of environmental factors on carbonate budget levels, we analyse live coral cover thresholds required for positive carbonate budgets from 10 to 20% (inclusive, at step 1%).

@@

## Reef carbonate budget
The number of years reefs maintain coral cover above a required carbonate budget level substantially decreases as the required carbonate budget threshold (as a percentage of reef area) increases. Reefs with greater median depth levels maintain coral cover above required carbonate budget levels for longer than reefs at shallow depths [@fig-access-esm1-5-depth-carbonate]. Reef median depth has a moderate positive correlation with the number of years a reef maintains coral cover above the required threshold [@fig-access-esm1-5-depth-carbonate]. Depth - carbonate budget maintainence correlation decreases as the required carbonate budget threshold increases. This is likely due to an increasing number of reefs being unable to maintain coral cover above threshold levels as the threshold increases. There are no clear relationships between reef total connectivity levels and the number of years reefs maintain coral cover above carbonate budget threshold levels [@sfig-access-esm1-5-conn-carbonate]. Reef log (natural) total connectivity strength has no correlation with the number of years a reef maintains cover above the required thresholds. Log connectivity strength - carbonate budget maintainence correlation slightly decreases as the required carbonate budget threshold increases.

![Number of years reef coral cover exceeds the carbonate budget threshold level from 2025 to 2099. The carbonate budget thresholds assessed range from 10 to 20% of total reef area. Bracketed x-axis values represent the Spearman's correlation between reef depth and the number of years cover exceeds the carbonate budget threshold. Point colour represents the median reef depth level in metres. Displays results from median timeseries using ACCESS-ESM1-5.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/depth_carbonate_scatter.png){#fig-access-esm1-5-depth-carbonate width=`{julia} carb_width`}

@@

@@

Across most bioregions there are not consistent differences between clusters in connectivity metrics such as source-to-sink connectivity ratio (@sfig-access-esm1-5-bioregion-logsosi; @sfig-access-cm2-bioregion-logsosi; @sfig-ec-earth3-veg-bioregion-logsosi; @sfig-gfdl-cm4-bioregion-logsosi; @sfig-noresm2-mm-bioregion-logsosi) and total connectivity strength (@sfig-access-esm1-5-bioregion-totalconnectivity; @sfig-access-cm2-bioregion-totalconnectivity; @sfig-ec-earth3-veg-bioregion-totalconnectivity; @sfig-gfdl-cm4-bioregion-totalconnectivity; @sfig-noresm2-mm-bioregion-totalconnectivity).

@@

Timeseries analysis has been used to identify three groups of reefs with similar declining (low), recovering (high), and intermediate (medium) coral cover trajectories throughout future scenarios. Under all GCMs most reefs suffer substantial declines in coral cover from 2025 levels. NorESM2-MM and ACCESS-ESM1-5 GCMs convey the most coral cover maintanance into the future. Differences between low, medium and high reef clusters in their timeseries are most notable under these two GCMs.

@@

Overall, our findings do not suggest that reefs in ‘high’ cover clusters have greater thermal tolerance than reefs in lower coral cover clusters. This is in contrast with literature suggesting corals have the capacity to adapt to increasing climate change pressures [@humanes2022; @lachs2024], including the investigation by Bairos-Novak et al. [-@bairos-novak2021] that revealed high heritability of bleaching and survival response traits used to inform the distributions of thermal tolerance traits used in this investigation. The lack of influence of thermal tolerance on reef outcomes is likely due to reef being impacted by overall similar levels of thermal stress, combined with the greater importance of factors such as reef depth and initial coral cover.

@@

Furthermore, our results suggest that the capacity of reefs to maintain healthy coral habitat by being above positive net carbonate budget levels is greater for reefs at deeper depths than shallow depths. While the maintanence of positive carbonate budget levels does not appear to be linked to reef connectivity metrics. The maintanence of positive carbonate budgets is important for reef resilience under climate change.

@@

ADRIA - CoralBlox has not been calibrated or validated at the GBR-wide scale. Accordingly, the purpose of this investigation is to assess the relative influence of reef characteristics on future resilience and highlight possible outcomes under highly uncertain future scenarios, rather than provide spatially explicit predictions of future resilience.

@@

![Number of years reef coral cover exceeds the carbonate budget threshold level from 2025 to 2099. The carbonate budget thresholds assessed range from 10 to 20% of total reef area. Bracketed x-axis values represent the Spearman's correlation between reef log (natural) total connectivity strength and the number of years cover exceeds the carbonate budget threshold. Point colour represents log (natural) scaled reef total connectivity strength. Displays results from median timeseries using ACCESS-ESM1-5.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/log_total_strength_carbonate_scatter.png){#sfig-access-esm1-5-conn-carbonate width=`{julia} carb_width`}

@@

![Log (natural) source – to – sink connectivity ratios for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines within each violin represent the median log (natural) source – to – sink ratio for each cluster and grey points represent the ratio value for each reef. Vertical grey lines represent a log (natural) source – to – sink ratio of 0 indicating equal outgoing and incoming connectivity levels. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/so_to_si_bioregion_violin.png){#sfig-access-esm1-5-bioregion-logsosi width=`{julia} violin_width`}

![Log (natural) total connectivity strength for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines within each violin represent the median log (natural) total connectivity strength for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/log_total_strength_bioregion_violin.png){#sfig-access-esm1-5-bioregion-totalconnectivity width=`{julia} violin_width`}

@@

![Number of years reef coral cover exceeds the carbonate budget threshold level from 2025 to 2099. The carbonate budget thresholds assessed range from 10 to 20% of total reef area. Bracketed x-axis values represent the Spearman's correlation between reef depth and the number of years cover exceeds the carbonate budget threshold. Point colour represents the median reef depth level in metres. Displays results from median timeseries using ACCESS-CM2.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-CM2/depth_carbonate_scatter.png){#sfig-access-cm2-depth-carbonate width=`{julia} carb_width`}

![Number of years reef coral cover exceeds the carbonate budget threshold level from 2025 to 2099. The carbonate budget thresholds assessed range from 10 to 20% of total reef area. Bracketed x-axis values represent the Spearman's correlation between reef log (natural) total connectivity strength and the number of years cover exceeds the carbonate budget threshold. Point colour represents log (natural) scaled reef total connectivity strength. Displays results from median timeseries using ACCESS-CM2.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-CM2/log_total_strength_carbonate_scatter.png){#sfig-access-cm2-conn-carbonate width=`{julia} carb_width`}
